# Reproducible Research: Peer Assessment 2
Author: Nitin sethi
Dec 20th 2016

## The effects of weather events on the United States public heath and economically

### Synopsis
The goal of this data analysis is to determine the more impactful types of
weather on the United States human population and economically between
1950-2011. I used injuries and fatalities to classify weather events that
affected population health most. I used property and crop damage to determine
which weather events were most economically impactful.

### Data Processing
The storm data is downloadable [Here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) [47Mb] and unzipped with bunzip2.

#### Basic Setup
```{r echo=TRUE}
echo=TRUE # Echo all code segments
options(scipen = 12)
library(R.utils)
library(ggplot2)
library(tidyr)
library(dplyr)
```

#### Downloand & unzip file
```{r cache = TRUE}
if(!file.exists('stormData.csv')) {
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile = "stormData.csv.bz2", method = 'curl')
  bunzip2("stormData.csv.bz2")
}
```

#### Load data
```{r cache = TRUE}
stormData <- read.csv('stormData.csv', stringsAsFactors = FALSE)
dim(stormData)
names(stormData)
```


#### Drop unused columns
```{r}
importantCols <- c("EVTYPE", "INJURIES", "FATALITIES", "PROPDMG", "PROPDMGEXP",
                   "CROPDMG", "CROPDMGEXP")
stormData <- stormData[importantCols]
dim(stormData)
```

#### Convert damage amounts into dollars
```{r}
convertMagnitude <- function(x) {
  if (x == '') {
    return(0)
  }

  switch(toupper(x),
          '?' = 0,
          '-' = NA,
          '+' = NA,
          'H' = 2,
          'K' = 3,
          'M' = 6,
          'B' = 9,
          x)
}

stormData$PROPDMGEXP <- sapply(as.character(stormData$PROPDMGEXP),
                               convertMagnitude)
stormData$PROPDMGEXP <- as.numeric(stormData$PROPDMGEXP)

stormData$PROPDMG <- stormData$PROPDMG * 10^stormData$PROPDMGEXP

stormData$CROPDMGEXP <- sapply(as.character(stormData$CROPDMGEXP),
                               convertMagnitude)
stormData$CROPDMGEXP <- as.numeric(stormData$CROPDMGEXP)

stormData$CROPDMG <- stormData$CROPDMG * 10^stormData$CROPDMGEXP


stormData$PROPDMGEXP <- NULL
stormData$CROPDMGEXP <- NULL


summary(stormData)
```

#### Group weather event types

There are many duplicated weather event types in the data, so I grouped similar
types of events and cleaned up type names.
```{r}
eventTypes <- levels(factor(stormData$EVTYPE))
length(eventTypes)
```

```{r}
stormData[grep('cold|freez|ice|winter|blizzard|snow|wintry|icy',
               stormData$EVTYPE,
               ignore.case=TRUE),]$EVTYPE = 'COLD WEATHER'
stormData[grep('heat|hot|warm|dry', stormData$EVTYPE,
               ignore.case=TRUE),]$EVTYPE = 'HEAT'
stormData[grep('wind|wnd', stormData$EVTYPE,
               ignore.case = TRUE),]$EVTYPE = 'WIND'
stormData[grep('tornado|waterspout|funnel', stormData$EVTYPE,
               ignore.case = TRUE),]$EVTYPE = 'TORNADO'
stormData[grep('flooding|flood|rain|wet', stormData$EVTYPE,
               ignore.case=TRUE),]$EVTYPE = 'FLOODING'
stormData[grep('dry|drought', stormData$EVTYPE,
               ignore.case=TRUE),]$EVTYPE = 'DROUGHT'
stormData[grep('fire', stormData$EVTYPE,
               ignore.case=TRUE),]$EVTYPE = 'WILD FIRE'
stormData[grep('hurricane|typhoon|surge', stormData$EVTYPE,
               ignore.case=TRUE),]$EVTYPE = 'HURRICANE'
stormData[grep('tropical', stormData$EVTYPE,
               ignore.case=TRUE),]$EVTYPE = 'TROPICAL STORM'
stormData[grep('surf|current|wave', stormData$EVTYPE,
               ignore.case=TRUE),]$EVTYPE = 'SURF'
stormData[grep('lightning', stormData$EVTYPE,
               ignore.case=TRUE),]$EVTYPE = 'LIGHTNING'
stormData[grep('hail', stormData$EVTYPE,
               ignore.case=TRUE),]$EVTYPE = 'HAIL'
eventTypes <- levels(factor(stormData$EVTYPE))
length(eventTypes)
```

### Data Analysis

I group the data by weather event types, sum the classifiers 
(FATALITIES, INJURIES) & (PROPDMG, CROPDMG), and sum the sum of the classifiers
which created a total impact for the weather type. I then order the weather 
types by their total impact descending.

#### Determine the most impactful storm events on the human population
```{r}
healthImpactData <- group_by(stormData, EVTYPE) %>%
    summarize(Deaths=sum(FATALITIES), Injuries=sum(INJURIES),
              Total=sum(INJURIES)+sum(FATALITIES)) %>%
    gather(Impact, ImpactTotal, -EVTYPE, -Total) %>%
    arrange(desc(Total))
healthImpactData <- healthImpactData[1:20,]
head(healthImpactData, n = 20)

```

#### Determine the storm events with the largest economic impact
```{r}
economicImpactData <- group_by(stormData, EVTYPE) %>%
    summarize(Property=sum(PROPDMG), Crop=sum(CROPDMG),
              Total=sum(PROPDMG)+sum(CROPDMG)) %>%
    gather(Impact, ImpactTotal, -EVTYPE, -Total) %>%
    arrange(desc(Total))
economicImpactData <- economicImpactData[1:20,]
head(economicImpactData, n = 20)
```

### Results

#### Impact on human population of the United States

The most impactful weather events that affect population health are tornadoes,
wind, heat, flooding, cold weather, and lightning. Tornadoes are a by far the
leading weather related cause for injuries and death in the United States from
1950-2011.

```{r}
# reorder event types for graph order
eventHealthTypes <- reorder(healthImpactData$EVTYPE, -healthImpactData$Total)

ggplot(healthImpactData, aes(x=eventHealthTypes, y=ImpactTotal/10^3,
                             fill=Impact)) +
  geom_bar(position="stack", stat="identity") +
  theme(axis.text.x=element_text(angle=90)) +
  labs(title = "Worst Weather related Fatalities and Injuries in the US 1950-2011",
       x = "Weather Event Type", y = "Total Impact (in thousands)")
```

#### Greatest economic consequences

The weather events that caused the greatest economic consequences are
hurricanes, cold winter weather, hail, drought, wild fires and tropical storms.
Hurricanes are the main weather related cause of economic destruction,
predominately property damage.  The leading cause of crop damage is drought, but
barely compares to the total economic impact by hurricanes.

```{r}
# reorder event types for graph order
eventEconTypes <- reorder(economicImpactData$EVTYPE, -economicImpactData$Total)

ggplot(economicImpactData, aes(x=eventEconTypes, y=ImpactTotal/10^9,
                               fill=Impact)) +
  geom_bar(position="stack", stat="identity") +
  theme(axis.text.x=element_text(angle=90)) +
  labs(title = "Worst Weather related Property & Crop Damage in the US 1950-2011",
       x = "Weather Event Type", y = "Total Damage (Billions of dollars)")
```
